# Cross-association {#sec-cross-correlation}

```{r}
#| label: setup
#| echo: false
#| results: asis
 
library(rebook)
chapterPreamble()
```

```{r}
#| label: load_pkg

library(mia)
```

Cross-association analysis is a straightforward approach that can
reveal strength and type of associations between data sets. For instance,
we can analyze if higher presence of a specific taxon relates to higher
levels of a biomolecule. Correlation analyses within dataset were already
discussed in [@sec-correlation].

As an example dataset, we use a data from the following publication:
[-@Hintikka2021] Xylo-Oligosaccharides in prevention of hepatic
steatosis and adipose tissue inflammation: associating taxonomic and
metabolomic patterns in fecal microbiota with biclustering.
In this study, mice were fed either with a high-fat or a low-fat diet,
and with or without prebiotics, for the purpose of studying whether prebiotics
attenuate the negative impact of a high-fat diet on health.

This example data can be loaded from the `mia` package. The data is
already in `MAE` format which is tailored for multi-assay analyses
(see [@sec-containers]). The dataset includes three different experiments:
microbial abundance data, metabolite concentrations, and data about
different biomarkers. If you like to construct the same data object from the
original files instead, [here]([@sec-loading-experimental-microbiome-data]) 
you can find help for importing data into an `SE`object.

```{r}
#| label: cross-association1

# Load the data
data(HintikkaXOData, package = "mia")
mae <- HintikkaXOData
```

```{r}
#| label: cross-association2

# Available alternative experiments
experiments(mae)
```

```{r}
#| label: cross-association3

# Microbiome data
getWithColData(mae, "microbiota")
```

```{r}
#| label: cross-association4

# Metabolite data
getWithColData(mae, "metabolites")
```

```{r}
#| label: cross-association5

# Biomarker data
getWithColData(mae, "biomarkers")
```

Next we can perform a cross-association analysis. Let us analyze if
individual bacteria genera are correlated with concentrations of
individual metabolites. This helps to answer the following question: "If
bacterium X is present, is the concentration of metabolite Y lower or higher"?

```{r}
#| label: cross-association6

# Agglomerate microbiome data at family level
mae[[1]] <- agglomerateByPrevalence(mae[[1]], rank = "Family", na.rm = TRUE)
# Does log10 transform for microbiome data
mae[[1]] <- transformAssay(mae[[1]], method = "log10", pseudocount = TRUE)

# Cross correlates data sets
res <- getCrossAssociation(
    mae,
    experiment1 = 1,
    experiment2 = 2,
    assay.type1 = "log10",
    assay.type2 = "nmr",
    method = "spearman",
    test.signif = TRUE,
    p.adj.threshold = NULL,
    cor.threshold = NULL,
    # Remove when mia is fixed
    mode = "matrix",
    sort = TRUE,
    show.warnings = FALSE)
```

Next, we create a heatmap depicting all cross-correlations between bacterial
genera and metabolite concentrations.

```{r}
#| label: cross-association7
#| fig-width: 10
#| fig-height: 8

library(ComplexHeatmap)
library(shadowtext)

# Function for marking significant correlations with "X"
add_signif <- function(j, i, x, y, width, height, fill) {
    # If the p-value is under threshold
    if( !is.na(res$p_adj[i, j]) & res$p_adj[i, j] < 0.05 ){
        # Print "X"
        grid.shadowtext(
            sprintf("%s", "X"), x, y, gp = gpar(fontsize = 8, col = "#f5f5f5"))
    }
}

# Create a heatmap
p <- Heatmap(res$cor,
    # Print values to cells
    cell_fun = add_signif,
    heatmap_legend_param = list(
        title = "correlation", legend_height = unit(5, "cm")),
    column_names_rot = 45
    )
p
```

::: {.callout-tip icon=false}
## Exercises

**Goal:** After completing these exercises, you should be able to calculate
associations between two omics.

**Exercise 1: Cross-association**

1. Load any of the example datasets in `MultiAssayExperiment` format mentioned
in [@sec-example-data].

2. Agglomerate microbiota data to prevalent phyla.

3. Apply the CLR transformation to microbiota data.

4. Apply the relative and log10 transformations to the other omic data.

5. Calculate correlations between the different omics data.

6. Visualize the results with a heatmap.

7. Are there taxonomic features that associate with the features of the other
omic? Do a literature search for the function of the features that show associations.
Do the associations make sense in light of the literature?

Useful functions:

`utils::data()`, `mia::agglomerateByPrevalence`, `mia::transformAssay()`,
`mia::getCrossAssociation()`, `ComplexHeatmap::Heatmap()`

:::
